Building Markov State Models for Periodically Driven Non-Equilibrium Systems

### 2014-09-22 writedown/period.noneq.msm/paper.tex
backup changes

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Spatial discretization: Master equation}
We will now introduce an appropriate spatial discretization of this kind of non-equilibirium MD -- this is done for reasons of simplicity only; we could completely avoid it for the price of more technical arguments.
For achieving this discretization, we 
introduce a partition of state space $\Omega$ into a finite number of disjoint
sets $\{ \Omega_1, \cdots, \Omega_N\}$ satisfying $\Omega = \cup_i \Omega_i$,
$\Omega_j\cap \Omega_j = \emptyset,\ \forall i\neq j$.
Utilizing the procedure described in Ref.~\cite{latorre2011structure} the original Fokker-Planck equation~(\ref{eq:disc-fp})
is discretized, resulting in a time-inhomogeneous Markov jump process in state
space $S = \{1, \cdots, N\}$ with time-dependent rate
matrix $\vect L(t) \in \mathbb R^{N\times N}$ satisfying
\begin{align}\label{eq:disc-4}
\sum\limits_{j=1}^N L_{ij}(t) & =  0\\ \label{eq:disc-5}
L_{ij}(t) & \ge  0, \quad i\not= j\\
L_{ij}(t) & =  L_{ij}(t+T)
\end{align}
for all real time $t\geq 0$.
Moreover, the rate matrix $L$ has the form $\vect L(t)=\vect L_0+E(t)\vect L_1$
where $E(t)$ is periodic with period $T>0$.
In analogy to \eqref{eq:disc-fp}, the Markov jump process generated by
$\vect L(t)$ transports probability distributions according to the associated Master equation
\begin{align}
  \label{eq:disc-master}
  \frac{d\vect p(t)}{dt} = \vect L^{\top}(t)\cdot \vect p(t)
\end{align}
where $\vect L^{\top}(t)$ denotes the matrix transpose of $\vect L(t)$, $\vect p(t)$ is an $N$-vector denoting the probability distribution on $S$ at time $t$, $p_i(t)$, for example, the probability to be in state $i$ (which corresponds to set $\Omega_i$) at time $t$.
As usual the properties (\ref{eq:disc-4}) and (\ref{eq:disc-5}) of
$\vect L(t)$ guarantee that the total probability mass is conserved,
i.e., if $p_i(0)\ge 0$ componentwise, then $p_i(t)\ge 0$ and $\sum_i
p_i(t) = \sum_ip_i(0)$.
% The solution of the master equation need no be
% periodic.
The temporal evolution of the probability distribution $\vect p(t)$ can be formally written
\begin{align}  \label{eq:disc-8}
\vect p(t)=\myphi(t)\vect p(0)
\end{align}
by using the
associated propagator matrix $\myphi(t)\in\mathbb R^{N\times N}$ that solves
\begin{align}
  \label{eq:disc-master-phi}
  \frac{d}{dt}\myphi(t) = \vect L^{\top}(t)\myphi(t), \quad \myphi(0) = \id .
\end{align}
Since the last equation can be considered column-wise, the propagator matrix inherits column-wise conservation properties:
$\Phi_{ij}(t) \ge  0$
and $\sum\limits_{i=1}^N \Phi_{ij}(t)  =  1$,
that is, $\myphi^{\top}(t)$ is a
stochastic matrix satisfying $\myphi^{\top}(t)\vect e=\vect e$
with $\vect e=(1,\ldots,1)^{\top}\in \mathbb R^N$.
Regarding these considerations, we find
\begin{align}
\label{eq:disc-10}  
\myphi_{ij}(t)=\prob\left(\vect X_t=i\mid \vect X_0=j \right),
\end{align}
where $\vect X_t$ denotes the Markov process generated by $\vect L(t)$.

The discretization sets that we used to go from $\vect x_t$ and $\ml(t)$ to $\vect X_t$
and $\vect L(t)$, respectively, can be assumed to provide an arbitrarily fine
partition of the original state space; then the transport properties
of $\vect L(t)$ are almost perfect approximations of the transport properties
of $\ml(t)$, in particular the approximation $p_i(t)\approx \prob(x_t\in \Omega_i)$ is almost perfect.

@@

Since $P$ is a stochastic matrix, its eigenvalues are contained in the unit circle in
the complex plane, i.e., each eigenvalue $\lambda$ (potentially coomplex-valued)
satisfies $|\lambda|\le 1$. Furthermore $\lambda=1$ is an
eigenvalue with right eigenvector $\vect e=(1,\ldots,1)^{\top}$ and a left eigenvector $\mymu$
satisfying
$\mymu^{\top}P=\mymu^{\top}$.
From now on, we assume $P$ to be irreducible and aperiodic such that the Perron Frobenius theorem the eigenvalue $\lambda=1$ is  non-negative componentwise, and unique (up to normalization $\sum_j\mu_j=1$). In this case $\mymu$ is the stationary measure in the sense that
$
 \mymu^{\top} P^m = \mymu^{\top},\ m\in\mathbb N
$,
% \begin{align}
% \label{eq:floq-14}  
% \end{align}
and (more precisely) the asymptotic evolution of an initial probability distribution $\vect p(t=0)$ by the process satisfies
$
\vect p^{\top}(0)\,P^m \to \mymu^{\top},\ m\to\infty,
$
% \begin{align}
% \label{eq:floq-15}  
% \end{align}
so that $\mymu$ can be seen as the quasi-stationary distribution of our non-stationary process.

@@

% \subsection{Building the Markov State Model from the Floquet transition matrix}
If the discretization cells $\Omega_i$, $i=1,\ldots,N$ form a fine
partition of the molecular state space, the Markov chain defined via
the transition matrix $P$
 is discrete in time, but in space it still is a fine-scale description of the transport properties of the
dynamics with a very large number $N$ of states.  Now we want to coarse our description much further by
constructing a Markov State Model (MSM) for $\vect P$ with $K\ll N$
\emph{macrostates} that should be the metastable states of the system: The resulting $K\times K$ MSM transition matrix $\hat{\vect P}$
then defines the coarse grained long term kinetics that shall
approximate the original long term kinetics well. 
The idea behind MSM building is that given the molecular system under consideration exhibits metastable conformations then it is usually possible to construct a
relatively small number of discrete sets --the metastable sets that form the so-called macrostates-- that
correctly describe the slow  dynamics, and 
in each set the fast dynamics relaxes on some timescales significantly shorter than the metastable timescales.
Then if the MSM dynamics reproduces the slow timescales and the corresponding transitions
of the original dynamics~\eqref{eq:disc-1},
the former is considered to be a good approximation of the later.

@@

Following the process described in Sec.~\ref{sec:build-msm}, we are able to build a 
three state MSM for the externally driven alanine dipeptide system,
where the quasi-stationary probability distribution $\mu$, the three core sets, and the forward and backward committors are
estimated as described above, and the MSM transition matrices $\hat P$, $\hat{M}$, and $\hat{T}$ are then evaluated using Eq.~\eqref{eq:msm-tmatrix-00}, \eqref{eq:msm-ht-1} and \eqref{eq:msm-hm-1}, correspondingly.
Alternatively, the MSM  transition matrix $\hat P$ is calculated directly from the NEMD trajectories using Eq.~\eqref{hatP_prob} (very good agreement with the one computed from the committors).
The leading eigenvalues of $P=\myphi^{\top}(T)$ are compared with those of $\hat P$ and $\hat{M}^{-1}\hat{T}$ in
Tab.~\ref{tab:tmp1}.
Without surprising, the two approaches for MSM building are consitent.
The MSM is able to accurately reproduce
the largest non-trivial eigenvalue, which means a precise reproduction
of the longest non-trivial implied time-scale. The accuracy of the second non-trivial
time-scale is not as good as the first, but is still acceptable.
The reason
for the lower accuracy may also be that
the corresponding time scale is 26.5~ps (calculated by $-T/\log(\lambda_2)$),
which is NOT significantly longer than the temporal resolution given by the period $T=10$~ps of the external driving force.
The difference between the eigenvalues of $\hat P$ and $\hat{M}^{-1}\hat T$
  can be taken as an indication for the non-Markovianity of the milestone process $\hat X_m$. This
  non-Markovianity seems to have stronger influence on the second non-trivial time-scale in comparison to the first one; this may be caused by shorter decorrelation timescales due to weaker metastability of the core sets involved.
Numerically the two MSM transition matrices are:
\[
\hat{M}^{-1}\hat{T}=\left(\begin{array}{ccc}
   0.860 &   0.134 &   0.007 \\
    0.190  &  0.775  &  0.034\\
    0.020   & 0.065 &   0.915
\end{array}
\right),\qquad \hat{P}=\left(\begin{array}{ccc}
   0.888 &   0.104 &   0.007 \\
    0.149  &  0.822  &  0.029\\
    0.021   & 0.056 &   0.923
\end{array}
\right),
\]
from which we see that  the left-handed alpha-helix conformations of the peptide exhibits the strongest metastability.

@@

\begin{table}
  \centering
  \caption{
    Comparison of second and third eigenvalues of $\vect P$ and $3\times 3$ MSM transition matrices $\hat{P}$ and $\hat{M}^{-1}\hat{T}$, respectively, from the two different MSM approaches.
  }
  \begin{tabular*}{0.5\textwidth}{@{\extracolsep{\fill}}c rrr}\hline\hline
      &  $\lambda_2$ & $\lambda_3$ & $\lambda_4$ \\\hline
    $P$                 &0.907  &0.686 & 0.553       \\
    $\hat P$     & 0.911  &0.723 & --       \\
    $\hat{M}^{-1}\hat{T}$    & 0.901  &0.649 & --       \\
    \hline\hline
  \end{tabular*}
  \label{tab:tmp1}
\end{table}

@@

In fact, $\hat{P}$ can be considered as the fingerprint of the long-term kinetics (cf. \cite{A19-39,PrinzKellerNoe_PCCP11_Perspective}) of alanine dipeptide in an oscillatory electric field.
In order to provide further validation of this statement, we study  time-dependent expectation values of
the form
\begin{align}
  \mathcal A(t) = \langle A(i)\rangle_t = \sum_{i\in S} A(i) p(i,t),
\end{align}
where $p(i,t) = p_i(t)=\prob(X_t=i)$ denotes the probability to be in set $\Omega_i$ at time $t$ as governed by the Master equation, and the observable $\mathcal A$ is spanned by the backward
committors, i.e.,
\begin{align}\label{A}
  A(i) = \sum_{j=1}^K \alpha_j q^\bwd_j(i).
\end{align}
Then
\begin{align}\nonumber
  \mathcal A(t) &=
  \sum_{i\in S} \sum_{j=1}^K \alpha_j q^\bwd_j(i)  p(i,t)=\sum_{i\in S} \sum_{j=1}^K \alpha_j \prob (\hat X_t = j \vert X_t = i) \prob (X_t = i) \\
  & =
  \sum_{i\in S} \sum_{j=1}^K \alpha_j \prob (\hat X_t = j ,X_t = i)  =
  \sum_{j=1}^K \alpha_j \prob (\hat X_t = j) 
 =
   \sum_{j=1}^K \alpha_j \,\hat p_j (t), \label{eq:num-28}
\end{align}
where the time-dependent probability $\hat{p}_j(t)$ of being assigned to MSM macrostate $j$ at time $t$ can be computed  by means of the MSM via simple matrix multiplications using (\ref{eq:num-29}).

@@

\recheck{Han: Here we have $p_j(t)$, $p(j,t)$ and $\hat p_j (0)$. That's a little confusing since I do not see a clear distinction between the former two. What can we do? What do you want to express?}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Spatial discretization: Master equation}
We will now introduce an appropriate spatial discretization of this kind of non-equilibirium MD -- this is done for reasons of simplicity only; we could completely avoid it for the price of more technical arguments.
For achieving this discretization, we 
introduce a partition of state space $\Omega$ into a finite number of disjoint
sets $\{ \Omega_1, \cdots, \Omega_N\}$ satisfying $\Omega = \cup_i \Omega_i$,
$\Omega_j\cap \Omega_j = \emptyset,\ \forall i\neq j$.
Utilizing the procedure described in Ref.~\cite{latorre2011structure} the original Fokker-Planck equation~(\ref{eq:disc-fp})
is discretized, resulting in a time-inhomogeneous Markov jump process in state
space $S = \{1, \cdots, N\}$ with time-dependent rate
matrix $\vect L(t) \in \mathbb R^{N\times N}$ satisfying
\begin{align}\label{eq:disc-4}
\sum\limits_{j=1}^N L_{ij}(t) & =  0\\ \label{eq:disc-5}
L_{ij}(t) & \ge  0, \quad i\not= j\\
L_{ij}(t) & =  L_{ij}(t+T)
\end{align}
for all real time $t\geq 0$.
Moreover, the rate matrix $L$ has the form $\vect L(t)=\vect L_0+E(t)\vect L_1$
where $E(t)$ is periodic with period $T>0$.
In analogy to \eqref{eq:disc-fp}, the Markov jump process generated by
$\vect L(t)$ transports probability distributions according to the associated Master equation
\begin{align}
  \label{eq:disc-master}
  \frac{d\vect p(t)}{dt} = \vect L^{\top}(t)\cdot \vect p(t)
\end{align}
where $\vect L^{\top}(t)$ denotes the matrix transpose of $\vect L(t)$, $\vect p(t)$ is an $N$-vector denoting the probability distribution on $S$ at time $t$, $p(i,t)$, for example, the probability to be in state $i$ (which corresponds to set $\Omega_i$) at time $t$.
As usual the properties (\ref{eq:disc-4}) and (\ref{eq:disc-5}) of
$\vect L(t)$ guarantee that the total probability mass is conserved,
i.e., if $p(i,0)\ge 0$ componentwise, then $p(i,t)\ge 0$ and $\sum_i
p(i,t) = \sum_ip(i,0)$.
% The solution of the master equation need no be
% periodic.
The temporal evolution of the probability distribution $\vect p(t)$ can be formally written
\begin{align}  \label{eq:disc-8}
\vect p(t)=\myphi(t)\vect p(0)
\end{align}
by using the
associated propagator matrix $\myphi(t)\in\mathbb R^{N\times N}$ that solves
\begin{align}
  \label{eq:disc-master-phi}
  \frac{d}{dt}\myphi(t) = \vect L^{\top}(t)\myphi(t), \quad \myphi(0) = \id .
\end{align}
Since the last equation can be considered column-wise, the propagator matrix inherits column-wise conservation properties:
$\Phi_{ij}(t) \ge  0$
and $\sum\limits_{i=1}^N \Phi_{ij}(t)  =  1$,
that is, $\myphi^{\top}(t)$ is a
stochastic matrix satisfying $\myphi^{\top}(t)\vect e=\vect e$
with $\vect e=(1,\ldots,1)^{\top}\in \mathbb R^N$.
Regarding these considerations, we find
\begin{align}
\label{eq:disc-10}  
\myphi_{ij}(t)=\prob\left(\vect X_t=i\mid \vect X_0=j \right),
\end{align}
where $\vect X_t$ denotes the Markov process generated by $\vect L(t)$.

The discretization sets that we used to go from $\vect x_t$ and $\ml(t)$ to $\vect X_t$
and $\vect L(t)$, respectively, can be assumed to provide an arbitrarily fine
partition of the original state space; then the transport properties
of $\vect L(t)$ are almost perfect approximations of the transport properties
of $\ml(t)$, in particular the approximation $p(i,t)\approx \prob(x_t\in \Omega_i)$ is almost perfect.

@@

Since $P$ is a stochastic matrix, its eigenvalues are contained in the unit circle in
the complex plane, i.e., each eigenvalue $\lambda$ (potentially coomplex-valued)
satisfies $|\lambda|\le 1$. Furthermore $\lambda=1$ is an
eigenvalue with right eigenvector $\vect e=(1,\ldots,1)^{\top}$ and a left eigenvector $\mymu$
satisfying
$\mymu^{\top}P=\mymu^{\top}$.
From now on, we assume $P$ to be irreducible and aperiodic such that the Perron Frobenius theorem the eigenvalue $\lambda=1$ is  non-negative componentwise, and unique (up to normalization $\sum_i\mu(i)=1$). In this case $\mymu$ is the stationary measure in the sense that
$
 \mymu^{\top} P^m = \mymu^{\top},\ m\in\mathbb N
$,
% \begin{align}
% \label{eq:floq-14}  
% \end{align}
and (more precisely) the asymptotic evolution of an initial probability distribution $\vect p(t=0)$ by the process satisfies
$
\vect p^{\top}(0)\,P^m \to \mymu^{\top},\ m\to\infty,
$
% \begin{align}
% \label{eq:floq-15}  
% \end{align}
so that $\mymu$ can be seen as the quasi-stationary distribution of our non-stationary process.

@@

% \subsection{Building the Markov State Model from the Floquet transition matrix}
If the discretization cells $\Omega_i$, $i=1,\ldots,N$ form a fine
partition of the molecular state space, the Markov chain defined via
the transition matrix $P$
 is discrete in time, but in space it still is a fine-scale description of the transport properties of the
dynamics with a very large number $N$ of states.  Now we want to coarse our description much further by
constructing a Markov State Model (MSM) for $\vect P$ with $K\ll N$
\emph{macrostates} that should be the metastable states of the system: The resulting $K\times K$ MSM transition matrix $\hat{\vect P}$
then defines the coarse grained long term kinetics that shall
approximate the original long term kinetics well. 
The idea behind MSM building is that given the molecular system under consideration exhibits metastable conformations then it is usually possible to construct a
relatively small number of discrete sets --the metastable sets that form the so-called macrostates-- that
correctly describe the slow  dynamics, and 
in each set the fast dynamics relaxes on some timescales significantly shorter than the metastable timescales.
Then if the MSM dynamics reproduces the slow timescales and the corresponding transitions
of the original dynamics~\eqref{eq:disc-1},
the former is considered to be a good approximation of the latter.

@@

Following the process described in Sec.~\ref{sec:build-msm}, we are able to build a 
three state MSM for the externally driven alanine dipeptide system,
where the quasi-stationary probability distribution $\mu$, the three core sets, and the forward and backward committors are
estimated as described above, and the MSM transition matrices $\hat P$, $\hat{M}$, and $\hat{T}$ are then evaluated using Eq.~\eqref{eq:msm-tmatrix-00}, \eqref{eq:msm-ht-1} and \eqref{eq:msm-hm-1}, correspondingly.
Alternatively, the MSM  transition matrix $\hat P$ is calculated directly from the NEMD trajectories using Eq.~\eqref{hatP_prob} (very good agreement with the one computed from the committors).
The leading eigenvalues of $P=\myphi^{\top}(T)$ are compared with those of $\hat P$ and $\hat{M}^{-1}\hat{T}$ in
Tab.~\ref{tab:tmp1}.
Without surprising, the two approaches for MSM building are consitent, \recheck{and the latter is slightly better}.
The MSM is able to accurately reproduce
the largest non-trivial eigenvalue, which means a precise reproduction
of the longest non-trivial implied time-scale. The accuracy of the second non-trivial
time-scale is not as good as the first, but is still acceptable.
The reason
for the lower accuracy may also be that
the corresponding time scale is 26.5~ps (calculated by $-T/\log(\lambda_2)$),
which is NOT significantly longer than the temporal resolution given by the period $T=10$~ps of the external driving force.
The difference between the eigenvalues of $\hat P$ and $\hat{M}^{-1}\hat T$
  can be taken as an indication for the non-Markovianity of the milestone process $\hat X_m$. This
  non-Markovianity seems to have stronger influence on the second non-trivial time-scale in comparison to the first one; this may be caused by shorter decorrelation timescales due to weaker metastability of the core sets involved.
Numerically the two MSM transition matrices are:
\[
\hat{M}^{-1}\hat{T}=\left(\begin{array}{ccc}
   0.860 &   0.134 &   0.007 \\
    0.190  &  0.775  &  0.034\\
    0.020   & 0.065 &   0.915
\end{array}
\right),\qquad \hat{P}=\left(\begin{array}{ccc}
   0.882 &   0.110 &   0.008 \\
    0.158  &  0.815  &  0.028\\
    0.023   & 0.055 &   0.922
\end{array}
\right),
\]
from which we see that  the left-handed alpha-helix conformations of the peptide exhibits the strongest metastability.

@@

\begin{table}
  \centering
  \caption{
    Comparison of second and third eigenvalues of $\vect P$ and $3\times 3$ MSM transition matrices $\hat{P}$ and $\hat{M}^{-1}\hat{T}$, respectively, from the two different MSM approaches.
  }
  \begin{tabular*}{0.5\textwidth}{@{\extracolsep{\fill}}c rrr}\hline\hline
      &  $\lambda_2$ & $\lambda_3$ & $\lambda_4$ \\\hline
    $P$                 &0.905  &0.668 & 0.551       \\
    $\hat P$     & 0.909  &0.710 & --       \\
    $\hat{M}^{-1}\hat{T}$    & 0.901  &0.649 & --       \\
    \hline\hline
  \end{tabular*}
  \label{tab:tmp1}
\end{table}

@@

In fact, $\hat{P}$ can be considered as the fingerprint of the long-term kinetics (cf. \cite{A19-39,PrinzKellerNoe_PCCP11_Perspective}) of alanine dipeptide in an oscillatory electric field.
In order to provide further validation of this statement, we study  time-dependent expectation values of
the form
\begin{align}
  \mathcal A(t) = \langle A(i)\rangle_t = \sum_{i\in S} A(i) p(i,t),
\end{align}
where $p(i,t)=\prob(X_t=i)$ is the probability to be in set $\Omega_i$ at time $t$ as governed by the Master equation, and the observable $\mathcal A$ is spanned by the backward
committors, i.e.,
\begin{align}\label{A}
  A(i) = \sum_{j=1}^K \alpha_j q^\bwd_j(i).
\end{align}
Then
\begin{align}\nonumber
  \mathcal A(t) &=
  \sum_{i\in S} \sum_{j=1}^K \alpha_j q^\bwd_j(i)  p(i,t)=\sum_{i\in S} \sum_{j=1}^K \alpha_j \prob (\hat X_t = j \vert X_t = i) \prob (X_t = i) \\
  & =
  \sum_{i\in S} \sum_{j=1}^K \alpha_j \prob (\hat X_t = j ,X_t = i)  =
  \sum_{j=1}^K \alpha_j \prob (\hat X_t = j) 
 =
   \sum_{j=1}^K \alpha_j \,\hat p_j (t), \label{eq:num-28}
\end{align}
where the time-dependent probability $\hat{p}_j(t)$ of being assigned to MSM macrostate $j$ at time $t$ can be computed  by means of the MSM via simple matrix multiplications using (\ref{eq:num-29}).

@@

% \recheck{Han: Here we have $p_j(t)$, $p(j,t)$ and $\hat p_j (0)$. That's a little confusing since I do not see a clear distinction between the former two. What can we do? What do you want to express?}
